---
title: "RMS Titanic"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

- The variables in this dataset are Class, Sex, Age, Survived, n (count).

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- There are 2201 passengers in this dataset. The wikipedia lists 2208 passengers.

- *Are there any differences?*

  - Yes. While the dataset lists 2201 passengers, the wikipedia page for the Titanic lists 2208 passengers, meaning there is a 7 person discrepancy

- *If yes, what might account for those differences?*

  - This is most likely due to Wikipedia being a "living document" that is collaborated on by many people. The original dataset is based on the count shortly after the titanic, but there may be families that came forward later or other information that has come forward since the original document. Wikipedia also says the passenger list "included some names of people who cancelled their trip at the last minute" and that there were some passengers who "traveled under aliases....and were therefore double-counted on the casualty lists".

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(x = Class, y = n, fill= Sex)) +
  geom_col(position = "dodge")
```

**Observations**:

- According to the data, in almost all classes more women survived, with the survival rates becoming closer with lower classes
- A key point here is that this trend is not seen with the crew.

Not an observation (?) but the use of n for the y axis means that for crew, we see few women surviving compared to men. There were far fewer women crew members, so there's a bit of a skew here that would not be seen if it was based on proportion or percent chance of survival.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge")
```

**Observations**:

- Both the male and female bars for 1st class and second class are at 1, which is 100% survival rate.

- For 3rd class and crew, the women had much higher survival proportions htan males, but are still overall lower than 1st and second class

- Is there anything *fishy* going on in your plot?

  - 100% bars for 1st and 2nd Class gives the idea that everyone survived
  - Warning message - removed 2 rows contianing missing values.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge") +
  facet_grid(~ Age)
```

**Observations**:

- Children had high survival in 1st and 2nd class (100%). In 3rd class, it was under 50% for female children and \~30% for male children.
- Across all categories, female survival rates were higher than men
- For adults, the survival rates decreased with class for the most part (exceptions for 2nd class males and crew)
- There were no crew children, which might explain the error message seen here and in q4
- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  - Based on this, It seems like because the data wasn't split into age, it was just putting the child data on top for the 1st and 2nd class. That resulted in it reading as 1st and 2nd class had 100% survival rates when it was just the 1st and 2nd class children that did.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
